---
title: "large-scale_PBMC_CRISPR_TRBJ1-1.2.6_mir223.Rmd"
date: "2026-06-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(plotROC)
library(pROC)
library(ggrepel)
library(FSA)
library(ComplexUpset)
```

## Calculate the Mean normalized fluorescence from the replicates
```{r}
rawData = read_delim(file = "large_scale_CRISPR_raw_data.tsv", delim = "\t")

rawData$`Reaction Conditions` = factor(rawData$`Reaction Conditions`, levels = c("PNK+/PolyA+/RT+","PNK+/PolyA+/RT-","Max Fluorescence in PCR buffer"))
rawData$Sample_Name = factor(rawData$Sample_Name, levels = c("Healthy PBMC","Non-IBC PBMC","IBC PBMC"))
rawData$`crRNA Target` = factor(rawData$`crRNA Target`, levels = c("TRBJ1-1","TRBJ1-2","TRBJ1-6","mir-223-3p"))
rawData$Replicate <- as.factor(rawData$Replicate)
rawData$`CRISPR Date` <- as.factor(rawData$`CRISPR Date`)

baselineData = rawData %>% 
  filter(Min <=0) %>% 
  select(`CRISPR ID`, Min, Fluorescence) %>% 
  pivot_wider(values_from = Fluorescence, names_from = Min) %>% 
  rowwise() %>%
  mutate(Baseline_mean = mean(c_across(`-10`:`0`))) %>% 
  select(`CRISPR ID`, Baseline_mean)

Fmax_TRBJ1_6_2025_05_02 = baselineData %>% 
  filter(`CRISPR ID` %in% c("B12_Fmax_2025-05-02")) %>% ## A12 and B12 are replicates, B12 is higher
  pull(Baseline_mean)

Fmax_TRBJ1_6_2025_05_19 = baselineData %>% 
  filter(`CRISPR ID` %in% c("B12_Fmax_2025-05-19")) %>% ## A12 and B12 are replicates, B12 is higher
  pull(Baseline_mean)

Fmax_mir223_2025_05_20 = baselineData %>% 
  filter(`CRISPR ID` %in% c("G12_Fmax_2025-05-20")) %>% ## G12 and H12 are replicates, G12 is higher
  pull(Baseline_mean)

Fmax_mir223_2025_05_22 = baselineData %>% 
  filter(`CRISPR ID` %in% c("G12_Fmax_2025-05-22")) %>% ## G12 and H12 are replicates, G12 is higher
  pull(Baseline_mean)
# ~~~~~~~~~~~~~~~~
Fmax_TRBJ1_1_2025_06_11 = baselineData %>% 
  filter(`CRISPR ID` %in% c("A12_Fmax_2025-06-11")) %>% ## A12 and B12 are replicates, A12 is higher
  pull(Baseline_mean)

Fmax_TRBJ1_1_2025_06_16 = baselineData %>% 
  filter(`CRISPR ID` %in% c("B12_Fmax_2025-06-16")) %>% ## A12 and B12 are replicates, B12 is higher
  pull(Baseline_mean)

Fmax_TRBJ1_2_2025_06_02 = baselineData %>% 
  filter(`CRISPR ID` %in% c("G12_Fmax_2025-06-02")) %>% ## G12 and H12 are replicates, G12 is higher
  pull(Baseline_mean)

Fmax_TRBJ1_2_2025_06_13 = baselineData %>% 
  filter(`CRISPR ID` %in% c("G12_Fmax_2025-06-13")) %>% ## G12 and H12 are replicates, G12 is higher
  pull(Baseline_mean)


# unicode for delta in R is \u0394, but it turned out I could just insert Δ directly
normalizedData = rawData %>% 
  left_join(baselineData) %>% 
  mutate("Fluorescence Recovery (ΔF%)" = case_when(`CRISPR Date` == "2025-05-02" ~ ((Fluorescence-Baseline_mean)/(Fmax_TRBJ1_6_2025_05_02-Baseline_mean))*100,
                                                    `CRISPR Date` == "2025-05-19" ~ ((Fluorescence-Baseline_mean)/(Fmax_TRBJ1_6_2025_05_19-Baseline_mean))*100,
                                                    `CRISPR Date` == "2025-05-20" ~ ((Fluorescence-Baseline_mean)/(Fmax_mir223_2025_05_20-Baseline_mean))*100,
                                                    `CRISPR Date` == "2025-05-22" ~ ((Fluorescence-Baseline_mean)/(Fmax_mir223_2025_05_22-Baseline_mean))*100,
                                                   `CRISPR Date` == "2025-06-11" ~ ((Fluorescence-Baseline_mean)/(Fmax_TRBJ1_1_2025_06_11-Baseline_mean))*100,
                                                   `CRISPR Date` == "2025-06-16" ~ ((Fluorescence-Baseline_mean)/(Fmax_TRBJ1_1_2025_06_16-Baseline_mean))*100,
                                                   `CRISPR Date` == "2025-06-02" ~ ((Fluorescence-Baseline_mean)/(Fmax_TRBJ1_2_2025_06_02-Baseline_mean))*100,
                                                   `CRISPR Date` == "2025-06-13" ~ ((Fluorescence-Baseline_mean)/(Fmax_TRBJ1_2_2025_06_13-Baseline_mean))*100)) 


meannormalizedData = normalizedData %>% 
  filter(`Sample Type`=="PBMC") %>%
  select(Min, `Donor ID`, Sample_Name,`Sample Type`,Replicate,`Reaction Conditions`,`crRNA Target`,`Fluorescence Recovery (ΔF%)`) %>% # move Sample, Sample_Number out since I omitted it above
  pivot_wider(values_from = `Fluorescence Recovery (ΔF%)`,names_from = Replicate) %>% 
  mutate("Mean Fluorescence Recovery (ΔF%)"= rowMeans(across(`A`:`B`))) #%>% 
  #select(-`A`,-`B`) # I'm keeping this in so that I can have one supplementary data file with the replicate values next to the mean values


write_delim(meannormalizedData, file = "large_scale_CRISPR_mean_normalized_data.tsv", delim = "\t")
```

## Figures 4A and 4B
```{r}

# meannormalizedData = read_delim(file = "large_scale_CRISPR_mean_normalized_data.tsv", delim = "\t")

# Figure 4A- got help from chatgpt to make the breaks consistent
meannormalizedData %>% 
  filter(
    `Sample Type` == "PBMC",
    `Reaction Conditions` == "PNK+/PolyA+/RT+",
    Min >= 0
  ) %>% 
  filter(
    (`crRNA Target` %in% c("TRBJ1-1", "TRBJ1-2") & Min < 65) |
    (`crRNA Target` %in% c("TRBJ1-6", "mir-223-3p"))
  ) %>% 
  ggplot(aes(
    x = Min,
    y = `Mean Fluorescence Recovery (ΔF%)`,
    color = Sample_Name,
    group = `Donor ID`
  )) +
  geom_line() +
  facet_grid(~`crRNA Target`, scales = "free_x") +
scale_x_continuous(
  breaks = function(x) {
    if (max(x) <= 65) c(0, 20, 40, 60)
    else c(0, 60, 120, 180)
  }
)  +
  scale_color_manual(values = c("#3b3b3b", "#4173C6", "#FF5733")) +
  theme_bw() +
  theme(
    legend.position = "top"
  )

# Figure 4B
meannormalizedData %>%
  filter(Min == 60) %>%
  filter(`Sample Type`=="PBMC" & `Reaction Conditions` == "PNK+/PolyA+/RT+") %>% 
  ggplot(aes(x = Sample_Name, y = `Mean Fluorescence Recovery (ΔF%)`, fill = Sample_Name))+
  geom_violin(aes(alpha = 0.5))+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5)+
  facet_wrap(.~`crRNA Target`, scales = "free_y", nrow = 1)+
  labs(title = graphName, subtitle = paste("Fluorescence at 60 min"))+
  theme(axis.text.x=element_text(angle=90,hjust = 1))+
  theme_bw()+
  scale_fill_manual(values = c("#3b3b3b","#4173C6","#FF5733"))+
  theme(axis.title.x=element_blank(),legend.position="top")

```

## Statistics at 60min timepoint
## TRBJ1-1: IBC vs Healhty and non-IBC significant
```{r}
# filter the data to just the timepoint used for the violin plot and ROC curve (if I change the timepoint I'll need to update this)
meannormalizedData_Min60_TRBJ1_1= meannormalizedData %>%
  filter(Min == 60) %>%
  filter(`Sample Type`=="PBMC" & `Reaction Conditions` == "PNK+/PolyA+/RT+" & `crRNA Target` == "TRBJ1-1")

# non-parametric test is needed since data aren't normal or equal variance, shows the same significant result
# part of base installation "stats" package
kruskal.test(`Mean Fluorescence Recovery (ΔF%)` ~ Sample_Name, data = meannormalizedData_Min60_TRBJ1_1)

# a follow up to Kruskal-Wallis test to get individual comparisons
# Perform Dunn's test
dunnTest(`Mean Fluorescence Recovery (ΔF%)` ~ Sample_Name, data = meannormalizedData_Min60_TRBJ1_1, method = "holm") # Holm is a good standard choice
```

## TRBJ1-2: IBC vs Healhty and non-IBC significant
```{r}

# filter the data to just the timepoint used for the violin plot and ROC curve (if I change the timepoint I'll need to update this)
meannormalizedData_Min60_TRBJ1_2= meannormalizedData %>%
  filter(Min == 60) %>%
  filter(`Sample Type`=="PBMC" & `Reaction Conditions` == "PNK+/PolyA+/RT+" & `crRNA Target` == "TRBJ1-2")

# non-parametric test is needed since data aren't normal or equal variance, shows the same significant result
# part of base installation "stats" package
kruskal.test(`Mean Fluorescence Recovery (ΔF%)` ~ Sample_Name, data = meannormalizedData_Min60_TRBJ1_2)

# a follow up to Kruskal-Wallis test to get individual comparisons
# Perform Dunn's test
dunnTest(`Mean Fluorescence Recovery (ΔF%)` ~ Sample_Name, data = meannormalizedData_Min60_TRBJ1_2, method = "holm") # Holm is a good standard choice
```

## TRBJ1-6: IBC vs Healhty and non-IBC significant
```{r}

# filter the data to just the timepoint used for the violin plot and ROC curve (if I change the timepoint I'll need to update this)
meannormalizedData_Min60_TRBJ1_6= meannormalizedData %>%
  filter(Min == 60) %>%
  filter(`Sample Type`=="PBMC" & `Reaction Conditions` == "PNK+/PolyA+/RT+" & `crRNA Target` == "TRBJ1-6")

# non-parametric test is needed since data aren't normal or equal variance, shows the same significant result
# part of base installation "stats" package
kruskal.test(`Mean Fluorescence Recovery (ΔF%)` ~ Sample_Name, data = meannormalizedData_Min60_TRBJ1_6)

# a follow up to Kruskal-Wallis test to get individual comparisons
# Perform Dunn's test
dunnTest(`Mean Fluorescence Recovery (ΔF%)` ~ Sample_Name, data = meannormalizedData_Min60_TRBJ1_6, method = "holm") # Holm is a good standard choice
```

## mir-223-3p = no comparisons significant
```{r}

# filter the data to just the timepoint used for the violin plot and ROC curve (if I change the timepoint I'll need to update this)
meannormalizedData_Min60_mir_223= meannormalizedData %>%
  filter(Min == 60) %>%
  filter(`Sample Type`=="PBMC" & `Reaction Conditions` == "PNK+/PolyA+/RT+" & `crRNA Target` == "mir-223-3p")

# non-parametric test is needed since data aren't normal or equal variance, shows the same significant result
# part of base installation "stats" package
kruskal.test(`Mean Fluorescence Recovery (ΔF%)` ~ Sample_Name, data = meannormalizedData_Min60_mir_223)

# a follow up to Kruskal-Wallis test to get individual comparisons
# Perform Dunn's test
dunnTest(`Mean Fluorescence Recovery (ΔF%)` ~ Sample_Name, data = meannormalizedData_Min60_mir_223, method = "holm") # Holm is a good standard choice
```


## determine AUC curves for IBC vs combined Healthy and non-IBC- Figure S6A
```{r}
IBC_vs_all = meannormalizedData %>% 
  filter(`Sample Type`=="PBMC" & `Reaction Conditions` == "PNK+/PolyA+/RT+" & `crRNA Target` != "mir-223-3p") %>%
  mutate(Disease = case_when(Sample_Name == "IBC PBMC" ~ 1,
                             Sample_Name == "Non-IBC PBMC" ~ 0,
                             Sample_Name == "Healthy PBMC" ~ 0))

# TRBJ1-1
TRBJ1_1_vsAll_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-1") %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 15)+#, labelround = 2
  facet_wrap(~ Min)

calc_auc(TRBJ1_1_vsAll_meannormalizedData)

# TRBJ1-2
TRBJ1_2_vsAll_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-2") %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 15)+#, labelround = 2
  facet_wrap(~ Min)

calc_auc(TRBJ1_2_vsAll_meannormalizedData)

# TRBJ1-6 AUC
TRBJ1_6_vsAll_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-6") %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 15)+#, labelround = 2
  facet_wrap(~ Min)

calc_auc(TRBJ1_6_vsAll_meannormalizedData)

# prep for Figure S6A
TRBJ1_1_AUC_vsAll = calc_auc(TRBJ1_1_vsAll_meannormalizedData) %>% 
  mutate(Target = "TRBJ1-1") %>% 
  select(Target, Min, AUC)

TRBJ1_2_AUC_vsAll = calc_auc(TRBJ1_2_vsAll_meannormalizedData) %>% 
  mutate(Target = "TRBJ1-2") %>% 
  select(Target, Min, AUC)

TRBJ1_6_AUC_vsAll = calc_auc(TRBJ1_6_vsAll_meannormalizedData) %>% 
  mutate(Target = "TRBJ1-6") %>% 
  select(Target, Min, AUC)

AUC_vsAll = TRBJ1_1_AUC_vsAll %>% 
  rbind(TRBJ1_2_AUC_vsAll) %>% 
  rbind(TRBJ1_6_AUC_vsAll)

AUC_vsAll$Target = factor(AUC_vsAll$Target, levels = c("TRBJ1-1","TRBJ1-2","TRBJ1-6"))

# Figure S6A
AUC_vsAll%>% 
  filter(Target %in% c("TRBJ1-1","TRBJ1-2") & Min < 65 | Target %in% c("TRBJ1-6")) %>%
  ggplot(aes(x = Min, y = AUC))+
  geom_point(shape = 1)+
  ylim(0,1)+
  theme_bw()+
  scale_x_continuous(
  breaks = function(x) {
    if (max(x) <= 65) c(0, 20, 40, 60)
    else c(0, 60, 120, 180)})+ # added with function from chatgpt
  labs(title = "IBC PBMC vs combined Healthy and non-IBC")+
  facet_wrap(.~ Target, scales = "free_x", nrow = 1)

```
## Figure 4C- ROC for IBC vs combined Healthy and Non-IBC
```{r}
Min60vsAll_meannormalizedData = IBC_vs_all %>% 
  filter(Min == 60) %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`, color = `crRNA Target`)) +
  geom_roc(n.cuts = 0)+
  style_roc()+
  scale_color_manual(values = c("purple","darkorange","darkred"))

TRBJ1_1_Min60_vsAll_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-1" & Min == "60") %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 0)

TRBJ1_2_Min60_vsAll_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-2" & Min == "60") %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 0)

TRBJ1_6_Min60_vsAll_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-6" & Min == "60") %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 0)

# Figure 4C
Min60vsAll_meannormalizedData+
  style_roc()+
  annotate("text", x = .25, y = .65, color = "purple",
            label = paste("TRBJ1-1 AUC =", round(calc_auc(TRBJ1_1_Min60_vsAll_meannormalizedData)$AUC, 2)))+ # 0.94
  annotate("text", x = .25, y = .7, color = "darkorange",
           label = paste("TRBJ1-2 AUC =", round(calc_auc(TRBJ1_2_Min60_vsAll_meannormalizedData)$AUC, 2)))+ # 0.95
  annotate("text", x = .25, y = .75, color = "darkred",
            label = paste("TRBJ1-6 AUC =", round(calc_auc(TRBJ1_6_Min60_vsAll_meannormalizedData)$AUC, 2)))+ # 0.98
  labs(title = "IBC PBMC vs combined Healthy and non-IBC at 60 min")+
  theme(legend.position="top")

```
## determine AUC curves for IBC vs Healthy- Figure S6B
```{r}
# TRBJ1-1
TRBJ1_1_vsHealthy_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-1" & Sample_Name != "Non-IBC PBMC") %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 15)+#, labelround = 2
  facet_wrap(~ Min)

calc_auc(TRBJ1_1_vsHealthy_meannormalizedData)

# TRBJ1-2
TRBJ1_2_vsHealthy_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-2" & Sample_Name != "Non-IBC PBMC") %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 15)+#, labelround = 2
  facet_wrap(~ Min)

calc_auc(TRBJ1_2_vsHealthy_meannormalizedData)

# TRBJ1-6 AUC
TRBJ1_6_vsHealthy_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-6" & Sample_Name != "Non-IBC PBMC") %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 15)+#, labelround = 2
  facet_wrap(~ Min)

calc_auc(TRBJ1_6_vsHealthy_meannormalizedData)


# prep for Figure S6B
TRBJ1_1_AUC_vsHealthy = calc_auc(TRBJ1_1_vsHealthy_meannormalizedData) %>% 
  mutate(Target = "TRBJ1-1") %>% 
  select(Target, Min, AUC)

TRBJ1_2_AUC_vsHealthy = calc_auc(TRBJ1_2_vsHealthy_meannormalizedData) %>% 
  mutate(Target = "TRBJ1-2") %>% 
  select(Target, Min, AUC)

TRBJ1_6_AUC_vsHealthy = calc_auc(TRBJ1_6_vsHealthy_meannormalizedData) %>% 
  mutate(Target = "TRBJ1-6") %>% 
  select(Target, Min, AUC)

AUC_vsHealthy = TRBJ1_1_AUC_vsHealthy %>% 
  rbind(TRBJ1_2_AUC_vsHealthy) %>% 
  rbind(TRBJ1_6_AUC_vsHealthy)

AUC_vsHealthy$Target = factor(AUC_vsHealthy$Target, levels = c("TRBJ1-1","TRBJ1-2","TRBJ1-6"))

# Figure S6B
AUC_vsHealthy%>% 
  filter(Target %in% c("TRBJ1-1","TRBJ1-2") & Min < 65 | Target %in% c("TRBJ1-6")) %>%
  ggplot(aes(x = Min, y = AUC))+
  geom_point(shape = 1)+
  ylim(0,1)+
  scale_x_continuous(
  breaks = function(x) {
    if (max(x) <= 65) c(0, 20, 40, 60)
    else c(0, 60, 120, 180)})+ # added with function from chatgpt
  theme_bw()+
  labs(title = "IBC PBMC vs Healthy")+
  facet_wrap(.~ Target, scales = "free_x", nrow = 1)

```

## Figure 4C- ROC for ROC for IBC vs Healthy
```{r}
Min60_vsHealthy_meannormalizedData = IBC_vs_all %>% 
  filter(Min == 60 & Sample_Name != "Non-IBC PBMC") %>%
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`, color = `crRNA Target`)) +
  geom_roc(n.cuts = 0)+
  style_roc()+
  scale_color_manual(values = c("purple","darkorange","darkred"))

TRBJ1_1_Min60_vsHealthy_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-1" & Min == "60" & Sample_Name != "Non-IBC PBMC") %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 0)

TRBJ1_2_Min60_vsHealthy_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-2" & Min == "60" & Sample_Name != "Non-IBC PBMC") %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 0)

TRBJ1_6_Min60_vsHealthy_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-6" & Min == "60" & Sample_Name != "Non-IBC PBMC") %>%
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 0)

# Figure 4C
Min60_vsHealthy_meannormalizedData+
  style_roc()+
  annotate("text", x = .25, y = .65, color = "purple",
            label = paste("TRBJ1-1 AUC =", round(calc_auc(TRBJ1_1_Min60_vsHealthy_meannormalizedData)$AUC, 2)))+# 0.97
  annotate("text", x = .25, y = .7, color = "darkorange",
           label = paste("TRBJ1-2 AUC =", round(calc_auc(TRBJ1_2_Min60_vsHealthy_meannormalizedData)$AUC, 2)))+# 0.97
  annotate("text", x = .25, y = .75, color = "darkred",
            label = paste("TRBJ1-6 AUC =", round(calc_auc(TRBJ1_6_Min60_vsHealthy_meannormalizedData)$AUC, 2)))+# 0.99
  labs(title = "IBC PBMC vs Healthy at 60 min")+
  theme(legend.position="top")

```

## determine AUC curves for IBC vs non-IBC- Figure S6C
```{r}
# TRBJ1-1
TRBJ1_1_vsNonIBC_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-1" & Sample_Name != "Healthy PBMC") %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 15)+
  facet_wrap(~ Min)

calc_auc(TRBJ1_1_vsNonIBC_meannormalizedData)

# TRBJ1-2
TRBJ1_2_vsNonIBC_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-2" & Sample_Name != "Healthy PBMC") %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 15)+#, labelround = 2
  facet_wrap(~ Min)

calc_auc(TRBJ1_2_vsNonIBC_meannormalizedData)

# TRBJ1-6 AUC
TRBJ1_6_vsNonIBC_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-6" & Sample_Name != "Healthy PBMC") %>% 
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 15)+#, labelround = 2
  facet_wrap(~ Min)

calc_auc(TRBJ1_6_vsNonIBC_meannormalizedData)

# prep for Figure S6C
TRBJ1_1_AUC_vsNonIBC = calc_auc(TRBJ1_1_vsNonIBC_meannormalizedData) %>% 
  mutate(Target = "TRBJ1-1") %>% 
  select(Target, Min, AUC)

TRBJ1_2_AUC_vsNonIBC = calc_auc(TRBJ1_2_vsNonIBC_meannormalizedData) %>% 
  mutate(Target = "TRBJ1-2") %>% 
  select(Target, Min, AUC)

TRBJ1_6_AUC_vsNonIBC = calc_auc(TRBJ1_6_vsNonIBC_meannormalizedData) %>% 
  mutate(Target = "TRBJ1-6") %>% 
  select(Target, Min, AUC)

AUC_vsNonIBC = TRBJ1_1_AUC_vsNonIBC %>% 
  rbind(TRBJ1_2_AUC_vsNonIBC) %>% 
  rbind(TRBJ1_6_AUC_vsNonIBC)

AUC_vsNonIBC$Target = factor(AUC_vsNonIBC$Target, levels = c("TRBJ1-1","TRBJ1-2","TRBJ1-6"))

# Figure S6C
AUC_vsNonIBC%>% 
  filter(Target %in% c("TRBJ1-1","TRBJ1-2") & Min < 65 | Target %in% c("TRBJ1-6")) %>%
  ggplot(aes(x = Min, y = AUC))+
  geom_point(shape = 1)+
  ylim(0,1)+
  scale_x_continuous(
  breaks = function(x) {
    if (max(x) <= 65) c(0, 20, 40, 60)
    else c(0, 60, 120, 180)})+ # added with function from chatgpt
  theme_bw()+
  labs(title = "IBC PBMC vs non-IBC")+
  facet_wrap(.~ Target, scales = "free_x", nrow = 1)

```

## Figure 4C- ROC for ROC for IBC vs Non-IBC
```{r}
Min60_vsNonIBC_meannormalizedData = IBC_vs_all %>% 
  filter(Min == 60 & Sample_Name != "Healthy PBMC") %>% # fixed this line, need to rerun the figure and put in the manuscript!!!!!
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`, color = `crRNA Target`)) +
  geom_roc(n.cuts = 0)+
  style_roc()+
  scale_color_manual(values = c("purple","darkorange","darkred"))

TRBJ1_1_Min60_vsNonIBC_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-1" & Min == "60" & Sample_Name != "Healthy PBMC") %>%
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 0)

TRBJ1_2_Min60_vsNonIBC_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-2" & Min == "60" & Sample_Name != "Healthy PBMC") %>%
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 0)

TRBJ1_6_Min60_vsNonIBC_meannormalizedData = IBC_vs_all %>% 
  filter(`crRNA Target` == "TRBJ1-6" & Min == "60" & Sample_Name != "Healthy PBMC") %>%
  ggplot(aes(d = Disease, m = `Mean Fluorescence Recovery (ΔF%)`)) +
  geom_roc(n.cuts = 0)

# Figure 4C
Min60_vsNonIBC_meannormalizedData+
  style_roc()+
  annotate("text", x = .25, y = .65, color = "purple",
            label = paste("AUC =", round(calc_auc(TRBJ1_1_Min60_vsNonIBC_meannormalizedData)$AUC, 2)))+ # 0.91
  annotate("text", x = .25, y = .7, color = "darkorange",
           label = paste("AUC =", round(calc_auc(TRBJ1_2_Min60_vsNonIBC_meannormalizedData)$AUC, 2)))+# 0.94
  annotate("text", x = .25, y = .75, color = "darkred",
            label = paste("AUC =", round(calc_auc(TRBJ1_6_Min60_vsNonIBC_meannormalizedData)$AUC, 2)))+# 0.97
  labs(title = "IBC PBMC vs Non-IBC at 60 min")+
  theme(legend.position="top")

```


## Final scatter plots (Figure 5)
* Extract the threshold for 10% FDR to add to the scatter plots (updated for confidence intervals on 2026-04-21)
```{r}
set.seed(123) # for confidence interval calculations

meannormalizedData_60 = meannormalizedData %>% 
  select(-A,-B) %>% 
  filter(Min == 60,
         `Reaction Conditions` == "PNK+/PolyA+/RT+") %>% 
  pivot_wider(names_from = `crRNA Target`,values_from = `Mean Fluorescence Recovery (ΔF%)`)

meannormalizedData_60_ROC = meannormalizedData %>% 
  filter(Min == 60) %>% 
  filter(`Sample Type`=="PBMC" & `Reaction Conditions` == "PNK+/PolyA+/RT+") %>%
  mutate(Disease = case_when(Sample_Name == "IBC PBMC" ~ 1,
                             Sample_Name == "Non-IBC PBMC" ~ 0,
                             Sample_Name == "Healthy PBMC" ~ 0))

# TRBJ1-1
TRBJ1_1_vsAll_meannormalizedData_60_ROC =  meannormalizedData_60_ROC %>% 
  filter(`crRNA Target` == "TRBJ1-1")

roc_obj_1.1 <- roc(response = TRBJ1_1_vsAll_meannormalizedData_60_ROC$Disease, predictor =TRBJ1_1_vsAll_meannormalizedData_60_ROC$`Mean Fluorescence Recovery (ΔF%)`)

# check AUC using pROC and compare to plotROC
auc(roc_obj_1.1)
# Area under the curve: 0.9354
# same as plotROC

# Extract all thresholds with corresponding FPR (1 - specificity)
roc_data_1.1 <- data.frame(
  threshold = roc_obj_1.1$thresholds,
  fpr = 1 - roc_obj_1.1$specificities,
  tpr = roc_obj_1.1$sensitivities
)
roc_data_1.1
# Find threshold closest to FPR = 0.10
target_1.1 <- roc_data_1.1[which.min(abs(roc_data_1.1$fpr - 0.10)), ]
target_1.1 # threshold = 91.10913, fpr = 91.10913, tpr = 91.10913
Threshold1.1 = unlist(target_1.1$threshold)

# TRBJ1-2
TRBJ1_2_vsAll_meannormalizedData_60_ROC =  meannormalizedData_60_ROC %>% 
  filter(`crRNA Target` == "TRBJ1-2")

roc_obj_1.2 <- roc(response = TRBJ1_2_vsAll_meannormalizedData_60_ROC$Disease, predictor =TRBJ1_2_vsAll_meannormalizedData_60_ROC$`Mean Fluorescence Recovery (ΔF%)`)

# check AUC using pROC and compare to plotROC
auc(roc_obj_1.2)
# Area under the curve: 0.9513
# same as plotROC

# Extract all thresholds with corresponding FPR (1 - specificity)
roc_data_1.2 <- data.frame(
  threshold = roc_obj_1.2$thresholds,
  fpr = 1 - roc_obj_1.2$specificities,
  tpr = roc_obj_1.2$sensitivities
)
roc_data_1.2
# Find threshold closest to FPR = 0.10
target_1.2 <- roc_data_1.2[which.min(abs(roc_data_1.2$fpr - 0.10)), ]
target_1.2 # threshold = 67.42161, fpr = 0.1016949, tpr = 0.875
Threshold1.2 = unlist(target_1.2$threshold)

# TRBJ1-6
TRBJ1_6_vsAll_meannormalizedData_60_ROC =  meannormalizedData_60_ROC %>% 
  filter(`crRNA Target` == "TRBJ1-6")

roc_obj_1.6 <- roc(response = TRBJ1_6_vsAll_meannormalizedData_60_ROC$Disease, predictor =TRBJ1_6_vsAll_meannormalizedData_60_ROC$`Mean Fluorescence Recovery (ΔF%)`)

# check AUC using pROC and compare to plotROC
auc(roc_obj_1.6)
# Area under the curve: 0.9767
# same as plotROC

# Extract all thresholds with corresponding FPR (1 - specificity)
roc_data_1.6 <- data.frame(
  threshold = roc_obj_1.6$thresholds,
  fpr = 1 - roc_obj_1.6$specificities,
  tpr = roc_obj_1.6$sensitivities
)
roc_data_1.6
# Find threshold closest to FPR = 0.10
target_1.6 <- roc_data_1.6[which.min(abs(roc_data_1.6$fpr - 0.10)), ]
target_1.6 # threshold = 16.0299, fpr = 0.1016949, tpr = 0.9375
Threshold1.6 = unlist(target_1.6$threshold)

# add 10% false positive rate calculated above to dotted lines on the scatter plots

# Figure 5B- scatterplot TRBJ1-6 vs TRBJ1-1
meannormalizedData_60 %>% 
  ggplot(aes(x = `TRBJ1-6`, y = `TRBJ1-1`, color = Sample_Name, group = `Donor ID`))+
  geom_point(shape = 1, size = 3)+
  labs(title = "FINAL PBMC Means", subtitle = paste("TRBJ1-6 vs TRBJ1-1 Mean Fluorescence at 60 min"))+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90,hjust = 1))+
  scale_color_manual(values = c("#3b3b3b","#4173C6","#FF5733"))+
  geom_hline(yintercept = Threshold1.1, linetype = "dashed", color = "gray") +  # TRBJ1-1 is y-axis, so threshold is horizontal
  geom_vline(xintercept = Threshold1.6, linetype = "dashed", color = "gray") +  # TRBJ1-6 is x-axis, so threshold is vertical
  geom_text_repel(
    data = subset(
      meannormalizedData_60,
      (Sample_Name %in% c("Healthy PBMC","Non-IBC PBMC") & (`TRBJ1-1` >= Threshold1.1 |`TRBJ1-6` >= Threshold1.6 ) |
        Sample_Name == "IBC PBMC" & (`TRBJ1-6` < Threshold1.6 | `TRBJ1-1` < Threshold1.1))
    ),
    aes(label = `Donor ID`),
    box.padding = 0.5,
    segment.color = "grey50",
    max.overlaps = Inf
  )

# Figure 5C- scatterplot TRBJ1-1 vs TRBJ1-2
meannormalizedData_60 %>% 
  ggplot(aes(x = `TRBJ1-1`, y = `TRBJ1-2`, color = Sample_Name, group = `Donor ID`))+
  geom_point(shape = 1, size = 3)+
  labs(title = "FINAL PBMC Means", subtitle = paste("TRBJ1-1 vs TRBJ1-2 Mean Fluorescence at 60 min"))+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90,hjust = 1))+
  scale_color_manual(values = c("#3b3b3b","#4173C6","#FF5733"))+
  geom_hline(yintercept = Threshold1.2, linetype = "dashed", color = "gray") +  # TRBJ1-2 is y-axis, so threshold is horizontal
  geom_vline(xintercept = Threshold1.1, linetype = "dashed", color = "gray") +  # TRBJ1-1 is x-axis, so threshold is vertical
  geom_text_repel(
    data = subset(
      meannormalizedData_60,
      (Sample_Name %in% c("Healthy PBMC","Non-IBC PBMC") & (`TRBJ1-1` >= Threshold1.1 |`TRBJ1-2` >= Threshold1.2 ) |
        Sample_Name == "IBC PBMC" & (`TRBJ1-2` < Threshold1.2 | `TRBJ1-1` < Threshold1.1))
    ),
    aes(label = `Donor ID`),
    box.padding = 0.5,
    segment.color = "grey50",
    max.overlaps = Inf
  )

# Figure 5D- scatterplot TRBJ1-6 vs TRBJ1-2
meannormalizedData_60 %>% 
  ggplot(aes(x = `TRBJ1-6`, y = `TRBJ1-2`, color = Sample_Name, group = `Donor ID`))+
  geom_point(shape = 1, size = 3)+
  labs(title = "FINAL PBMC Means", subtitle = paste("TRBJ1-6 vs TRBJ1-2 Mean Fluorescence at 60 min"))+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90,hjust = 1))+
  scale_color_manual(values = c("#3b3b3b","#4173C6","#FF5733"))+
  geom_hline(yintercept = Threshold1.2, linetype = "dashed", color = "gray") +  # TRBJ1-2 is y-axis, so threshold is horizontal
  geom_vline(xintercept = Threshold1.6, linetype = "dashed", color = "gray") +  # TRBJ1-6 is x-axis, so threshold is vertical
  geom_text_repel(
    data = subset(
      meannormalizedData_60,
      (Sample_Name %in% c("Healthy PBMC","Non-IBC PBMC") & (`TRBJ1-2` >= Threshold1.2 |`TRBJ1-6` >= Threshold1.6 ) |
        Sample_Name == "IBC PBMC" & (`TRBJ1-6` < Threshold1.6 | `TRBJ1-2` < Threshold1.2))
    ),
    aes(label = `Donor ID`),
    box.padding = 0.5,
    segment.color = "grey50",
    max.overlaps = Inf
  )
```

## Upset plot (Figure 5A)
```{r}

meannormalizedData_60_for_upset_labeled = meannormalizedData_60 %>% 
  mutate("Detection TRBJ1-1" = case_when(`TRBJ1-1` > Threshold1.1 ~ TRUE,
                                       TRUE ~ FALSE),
         "Detection TRBJ1-2" = case_when(`TRBJ1-2` > Threshold1.2 ~ TRUE,
                                       TRUE ~ FALSE),
         "Detection TRBJ1-6" = case_when(`TRBJ1-6` > Threshold1.6 ~ TRUE,
                                       TRUE ~ FALSE)) %>% 
  select(`Donor ID`, Sample_Name, `Detection TRBJ1-1`, `Detection TRBJ1-2`, `Detection TRBJ1-6`) %>% 
  mutate(
    "intersection" = case_when(
      `Detection TRBJ1-1` == FALSE & `Detection TRBJ1-2` == FALSE & `Detection TRBJ1-6` == FALSE ~ "000",
      `Detection TRBJ1-1` == TRUE & `Detection TRBJ1-2` == TRUE & `Detection TRBJ1-6` == TRUE ~ "126",
      `Detection TRBJ1-1` == TRUE & `Detection TRBJ1-2` == FALSE & `Detection TRBJ1-6` == TRUE ~ "106",
      `Detection TRBJ1-1` == FALSE & `Detection TRBJ1-2` == FALSE & `Detection TRBJ1-6` == TRUE ~ "006",
      `Detection TRBJ1-1` == TRUE & `Detection TRBJ1-2` == TRUE & `Detection TRBJ1-6` == FALSE ~ "120",
      `Detection TRBJ1-1` == FALSE & `Detection TRBJ1-2` == TRUE & `Detection TRBJ1-6` == TRUE ~ "026",
      `Detection TRBJ1-1` == FALSE & `Detection TRBJ1-2` == TRUE & `Detection TRBJ1-6` == FALSE ~ "020",
      `Detection TRBJ1-1` == TRUE & `Detection TRBJ1-2` == FALSE & `Detection TRBJ1-6` == FALSE ~ "100",
      TRUE ~ "other"
    )) %>% 
  mutate(for_graph = case_when(
    intersection == "126" & Sample_Name %in% c("Healthy PBMC", "Non-IBC PBMC") ~ `Donor ID`,
    !(intersection %in% c("000","126")) ~ `Donor ID`
   ))

# to get the count for each sample type in each intersection category
meannormalizedData_60_for_upset_labeled %>% 
  group_by(Sample_Name,intersection) %>% 
  summarize(count = n())


# for some reason I needed to list the intersect as 6,2,1 in order for the resulting graph to have 1,2,6...
## here's the graph with the samples labeled, they are overlapped, but I can move them in illustrator
## for some reason the labels are duplicated when I export, but I have to manually adjust them anyway
upset(meannormalizedData_60_for_upset_labeled,
      intersect = c("Detection TRBJ1-6","Detection TRBJ1-2","Detection TRBJ1-1"),
      sort_sets = FALSE,
      base_annotations = list(
      'Intersection size' = intersection_size(aes(fill = Sample_Name),) +
  scale_fill_manual(values = c(
    "Healthy PBMC" = "#3b3b3b",
    "Non-IBC PBMC" = "#4173C6",
    "IBC PBMC" = "#FF5733"))
    ),
  annotations = list(
    "Samples" = ggplot(mapping = aes(x = intersection, y= 0, label = for_graph)) +
      geom_text(stat = "identity", size = 2)
  )
  )
```