---
title: "CRISPR_model.Rmd"
date: "2026-06-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
library(ComplexUpset)
library(ggrepel)
library(FSA)
library(caret)
library(car)
library(dplyr)
library(pROC)
library(tibble)
```

## Model (used chatgpt)
## this one combines the Healthy and non-IBC into outcome other for combined comparison to IBC
```{r}
Final_mean_PBMC_CRISPR_60 = read_delim(file = "Final_mean_PBMC_CRISPR_60min.tsv", delim = "\t")

Final_mean_PBMC_CRISPR_60_wide <- Final_mean_PBMC_CRISPR_60 %>%
  pivot_wider(
    names_from = `crRNA Target`,
    values_from = `Mean Fluorescence Recovery (ΔF%)`
  ) %>% 
  mutate(outcome = ifelse(`Sample Type` == "IBC", "IBC", "Other"),
         outcome = factor(outcome, levels = c("Other", "IBC")))

Final_mean_PBMC_CRISPR_60_wide$`Sample Type` <- factor(Final_mean_PBMC_CRISPR_60_wide$`Sample Type`,
                               levels = c("Healthy", "non-IBC", "IBC"))
```

## PCA (Figure 6A)
```{r}
# coded with chatgpt
# categorize the false positives false negatives based on Upset plot in Fig 
labels = c("H04","N39","N25","I14","I23","I25","N11","N16","I02","H23","N17","I32","N23","I11","N19")

pca_data <- Final_mean_PBMC_CRISPR_60_wide %>%
  select(`TRBJ1-1`,`TRBJ1-2`,`TRBJ1-6`) %>%
  scale()

pca <- prcomp(pca_data, scale. = TRUE)

percent_var <- (pca$sdev^2 / sum(pca$sdev^2)) * 100

pc1 <- round(percent_var[1], 1)
pc2 <- round(percent_var[2], 1)

xlab <- paste0("PC1 (", pc1, "%)")
ylab <- paste0("PC2 (", pc2, "%)")

pca_df <- as.data.frame(pca$x) %>%
  mutate(`Sample Type` = Final_mean_PBMC_CRISPR_60_wide$`Sample Type`,
         `Donor ID` = Final_mean_PBMC_CRISPR_60_wide$`Donor ID`,
         to_label = case_when(`Donor ID` %in% labels ~ `Donor ID`,
                               TRUE ~ NA))

ggplot(pca_df, aes(PC1, PC2, color = `Sample Type`, label = to_label)) +
  geom_point(size = 3, alpha = 0.75) +
  stat_ellipse() +
  theme_minimal()+
  scale_color_manual(values = c(
    "Healthy" = "#3b3b3b",
    "non-IBC" = "#4173C6",
    "IBC" = "#FF5733"))+
  theme_bw()+
  geom_text_repel(aes(label = to_label),
                  box.padding = 0.5,
    segment.color = "grey50",
    max.overlaps = Inf
  )+
  xlab(paste0("PC1 (", round(percent_var[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(percent_var[2], 1), "%)"))

```

## create a weighted combination of biomarkers based on the combined model (Figure 6B)
```{r}
set.seed(123)
Final_mean_PBMC_CRISPR_60_wide$outcome <- factor(
  Final_mean_PBMC_CRISPR_60_wide$outcome,
  levels = c("Other", "IBC")
)
model_combined <- glm(outcome ~ `TRBJ1-1` + `TRBJ1-2` + `TRBJ1-6`,
                      data = Final_mean_PBMC_CRISPR_60_wide,
                      family = binomial)

summary(model_combined)

Final_mean_PBMC_CRISPR_60_wide$biomarker_score <- predict(model_combined, type = "link")

ggplot(Final_mean_PBMC_CRISPR_60_wide, aes(x = `Sample Type`, y = biomarker_score, fill = `Sample Type`)) +
  geom_violin(alpha = 0.5, position = "identity")+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5)+
  scale_fill_manual(values = c(
    "Healthy" = "#3b3b3b",
    "non-IBC" = "#4173C6",
    "IBC" = "#FF5733"))+
  ylab("Combined Biomarker Score")+
  theme_bw()

```

## stats for combined biomarker score: IBC vs Healhty and non-IBC significant
```{r}

# the data to use
Final_mean_PBMC_CRISPR_60_wide

# non-parametric test is needed since data aren't normal or equal variance, shows the same significant result
# part of base installation "stats" package
kruskal.test(biomarker_score ~ `Sample Type`, data = Final_mean_PBMC_CRISPR_60_wide)

# a follow up to Kruskal-Wallis test to get individual comparisons
# Perform Dunn's test
dunnTest(biomarker_score ~ `Sample Type`, data = Final_mean_PBMC_CRISPR_60_wide, method = "holm") # Holm is a good standard choice
```

## cross-validation
```{r}
# all the variables are below 5, suggesting the biomarkers aren't colinear
vif(model_combined)

# `TRBJ1-1` `TRBJ1-2` `TRBJ1-6` 
 # 1.748886  2.127203  1.732564 

```

```{r}
set.seed(123)

Final_mean_PBMC_CRISPR_60_wide$outcome <- factor(
  Final_mean_PBMC_CRISPR_60_wide$outcome,
  levels = c("Other", "IBC")
)

train_control <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

cv_model <- train(
  outcome ~ `TRBJ1-1` + `TRBJ1-2` + `TRBJ1-6`,
  data = Final_mean_PBMC_CRISPR_60_wide,
  method = "glm",
  family = "binomial",
  metric = "ROC",
  trControl = train_control
)

cv_model
  # ROC       Sens       Spec     
  # 0.963254  0.9469697  0.8485714

```

## ROC characteristics for the cross-validated model with composite biomarker (Table 3)
```{r}


set.seed(123)

# =========================
# 1. Mean CV metrics
#    (caret default threshold = 0.5)
# =========================

cv_summary <- cv_model$resample %>%
  summarise(
    mean_auc  = mean(ROC),
    sd_auc    = sd(ROC),

    mean_sens = mean(Sens),
    sd_sens   = sd(Sens),

    mean_spec = mean(Spec),
    sd_spec   = sd(Spec),

    mean_fpr  = mean(1 - Spec),
    sd_fpr    = sd(1 - Spec)
  )

# =========================
# 2. Aggregate repeated CV predictions
# =========================

pooled_preds <- cv_model$pred %>%
  group_by(rowIndex) %>%
  summarise(
    obs = dplyr::first(obs),
    prob_IBC = mean(IBC)
  ) %>%
  ungroup()

# Ensure factor ordering
pooled_preds$obs <- factor(
  pooled_preds$obs,
  levels = c("Other", "IBC")
)

# =========================
# 3. ROC curve from pooled predictions
# =========================

roc_obj <- roc(
  response = pooled_preds$obs,
  predictor = pooled_preds$prob_IBC,
  levels = c("Other", "IBC"),
  direction = "<",
  quiet = TRUE
)

# =========================
# 4. AUC + bootstrap CI
# =========================

auc_val <- as.numeric(auc(roc_obj))

auc_ci <- ci.auc(
  roc_obj,
  method = "bootstrap",
  boot.n = 5000
)

# =========================
# 5. Select threshold
#    Closest to 10% FPR
# =========================

roc_df <- coords(
  roc_obj,
  x = "all",
  ret = c("threshold", "sensitivity", "specificity"),
  transpose = FALSE
)

roc_df <- roc_df %>%
  mutate(
    fpr = 1 - specificity
  )

best_row <- roc_df %>%
  dplyr::slice(which.min(abs(fpr - 0.10)))

threshold_prob <- best_row$threshold

sensitivity <- best_row$sensitivity
specificity <- best_row$specificity
fpr         <- best_row$fpr

# =========================
# 6. Publication-style table
# =========================

performance_table <- tibble(
  Metric = c(
    "Cross-validated AUC ROC",
    "Mean CV sensitivity (threshold = 0.5)",
    "Mean CV specificity (threshold = 0.5)",
    "Mean CV false positive rate (threshold = 0.5)",
    "Threshold selection criterion",
    "Selected probability threshold",
    "Sensitivity at selected threshold",
    "Specificity at selected threshold",
    "False positive rate at selected threshold"
  ),

  Value = c(

    sprintf(
      "%.3f (95%% CI %.3f–%.3f)",
      auc_val,
      auc_ci[1],
      auc_ci[3]
    ),

    sprintf(
      "%.1f%% ± %.1f%%",
      cv_summary$mean_sens * 100,
      cv_summary$sd_sens * 100
    ),

    sprintf(
      "%.1f%% ± %.1f%%",
      cv_summary$mean_spec * 100,
      cv_summary$sd_spec * 100
    ),

    sprintf(
      "%.1f%% ± %.1f%%",
      cv_summary$mean_fpr * 100,
      cv_summary$sd_fpr * 100
    ),

    "Closest to 10% FPR",

    sprintf("%.3f", threshold_prob),

    sprintf("%.1f%%", sensitivity * 100),

    sprintf("%.1f%%", specificity * 100),

    sprintf("%.1f%%", fpr * 100)
  )
)

performance_table
```



