---
title: "TGIRTseq_IBC_TRBJ1-1.2.6_Plasma_PBMC_PBMC-filtering.Rmd"
date: "2026-06-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tools)
library(tidyverse)
```
## This is the bash script I used to generated the <= 30 nt filtered datasets from PBMCs

```{bash, eval=FALSE}

# Filtered_30nt_or_less_Read1.sh 

#!/bin/bash
# requires samtools v1.20 for filtering capabilities
# this script takes a filename list as the input for samtools view and applying a filter to only keep the read1 reads (-f 67) that are <=30nt (-e 'length(seq)<30') and when the sequence length is equal to the transcript length (+/- direction doesn't matter)
# include the header in the output (-h) and specify the bam output file with samtools flag (-o)
# after making the size-selected bam file, runs bedtools coverage which is used to make the counts file

for inputFile in "$@"; do
        outputFile1=${inputFile/.bam/_30nt_filter.bam}
        outputFile2=${outputFile1/.bam/.bedcoverage}

        samtools view -h -f 67 -e 'length(seq)<=30 && (length(seq)==tlen || length(seq)==-tlen)' -o Filtered_30nt_or_less_Read1/${outputFile1} ${inputFile}

        bedtools coverage -a genes.bed -b Filtered_30nt_or_less_Read1/${outputFile1} > Filtered_30nt_or_less_Read1/bedtools_coverage/${outputFile2}
done
```

## Make the counts file from the bedcoverage file
```{r, eval = FALSE}
metadata_Wylie = read.delim(file = "IBC_DESeq2.metadata.tsv", sep = '\t') %>% 
  filter(Sample.Type == "PBMC")

IBC_BC_Files = list.files(path = "Filtered_30nt_or_less_Read1/bedtools_coverage/", pattern = ".bedcoverage", full.names = TRUE)

IBC_BC_Files_short_path = list.files(path = "Filtered_30nt_or_less_Read1/bedtools_coverage/", pattern = ".bedcoverage", full.names = FALSE)

header_coverage = c("chr","posn_start","posn_end","name","score","strand","gene","gene_match","counts","bases_covered","gene_length","fraction_bases_covered")

IBC_BC_df = tibble()

for (i in 1:length(IBC_BC_Files)) {
  indiv_data = read.delim(file = IBC_BC_Files[i], sep = "\t", header = FALSE)
  colnames(indiv_data) <- header_coverage
  indiv_data = indiv_data %>%
  mutate(filename = IBC_BC_Files_short_path[i]) %>%
  mutate(sample = str_extract(string = filename, pattern = "^[A-Za-z0-9]*"))
  IBC_BC_df = rbind(IBC_BC_df, indiv_data)
}

IBC_BC_df = IBC_BC_df %>%   
  left_join(metadata_Wylie, by = c("sample" = "Filename"))

# omit these genes with duplicate entries (it's tRNA gene names, but these libraries have already filtered out those reads in a prior mapping step in the pipeline)
duplicate_names = IBC_BC_df %>% 
  dplyr::group_by(name, DESeq2_sample_name) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop") %>%
  dplyr::filter(n > 1L) %>% 
  select(name) %>% 
  distinct() %>% 
  unlist(use.names = FALSE)

IBC_BC_df_unique = IBC_BC_df %>% 
  filter(!(name %in% duplicate_names))%>% 
  select(name,counts,DESeq2_sample_name) %>% 
  pivot_wider(id_cols = name, names_from = DESeq2_sample_name, values_from = counts)

# make the first column into a rownames column, so that there will be the same number of columns as there are rows in the metadata table
counts = IBC_BC_df_unique %>% 
  column_to_rownames(var = "name")

write.table(counts, file="30filteredPBMC.rawcounts", sep="\t", quote=F, col.names=NA) # want to keep rownames so use write.table

```


## DESeq2 data generated for the <= 30 nt filtered datasets from PBMCs
* the unfiltered PBMC and Plasma datasets were in the Supplemental File for Wylie et al. (2026)

```{r, eval = FALSE}
counts_30filter = read.delim(file="30filteredPBMC.rawcounts.tsv", sep = "\t", row.names = 1)


metadata_Wylie = read.delim(file = "IBC_DESeq2.metadata.tsv", sep = '\t') %>% 
  filter(Sample.Type == "PBMC")

 all(rownames(metadata_Wylie) == colnames(counts_30filter))

dds_30filter = DESeqDataSetFromMatrix(countData = counts_30filter,
                              colData = metadata_Wylie,
                              design= ~ Disease.category)

dds_30filter

dds_30filter = DESeq(dds_30filter)

# statistics for IBC versus Healthy for <=30nt filtered PBMC datasets
res_IBC_v_H = results(dds_30filter, contrast = c("Disease.category","IBC","Healthy"), tidy = TRUE)

# statistics for IBC versus non-IBC for <=30nt filtered PBMC datasets
res_IBC_v_N = results(dds_30filter, contrast = c("Disease.category","IBC","non-IBC"), tidy = TRUE)

write_delim(res_IBC_v_H, file = "PBMC_IBCvsH_genome_filtered_30_stats.tsv", delim = "\t")
write_delim(res_IBC_v_N, file = "PBMC_IBCvsN_genome_filtered_30_stats.tsv", delim = "\t")

# export normalized counts for analysis
normalized_counts = counts(dds_30filter, normalized=TRUE) %>% 
  as.data.frame() %>% 
  rownames_to_column("ID")

write.table(normalized_counts, file="IBC_PBMC_filtered30nt_DESeq2_normalized_counts.tsv", sep="\t", quote=F, col.names=NA)

```


## All statistics compiled for filtered and unfiltered datasets

```{r}
# files generated from the supplementary file for plasma and PBMC

TRBJ1 = c("TRBJ1_1","TRBJ1_2","TRBJ1_6","TRBJ1-1","TRBJ1-2","TRBJ1-6")

#### PBMC unfiltered stats from original paper ###########################
# nothing significant between IBC and Healthy PBMC without read filtering <=30nt
read_delim(file = "PBMC_IBCvsH_genome_stats.tsv", delim = "\t") %>%
  filter(ID %in% TRBJ1)

# significant between IBC and non-IBC PBMC without read filtering <=30nt: TRBJ1-6 (not 1-1, 1-2 is close but not significant)
read_delim(file = "PBMC_IBCvsN_genome_stats.tsv", delim = "\t") %>%
  filter(ID %in% TRBJ1)


#### PBMC filtered stats from this manuscript ###########################
# significant between IBC and Healthy PBMC without read filtering <=30nt: TRBJ1-6 (not 1-1, 1-2 is close but not significant)
read_delim(file = "PBMC_IBCvsH_genome_filtered_30_stats.tsv", delim = "\t") %>%
  filter(row %in% TRBJ1)

# significant between IBC and non-IBC PBMC without read filtering <=30nt: TRBJ1-6 (not 1-1, 1-2 is close but not significant)
read_delim(file = "PBMC_IBCvsN_genome_filtered_30_stats.tsv", delim = "\t") %>%
  filter(row %in% TRBJ1)


#### plasma stats from original paper ###########################
# significant between IBC and Healthy: TRBJ1-1,1-2,1-6 (not 1-5)
read_delim(file = "Plasma_IBCvsH_genome_stats.tsv", delim = "\t") %>%
  filter(ID %in% TRBJ1)

# significant between IBC and non-IBC: TRBJ1-1,1-5,1-6 (not 1-2)
read_delim(file = "Plasma_IBCvsN_genome_stats.tsv", delim = "\t") %>%
  filter(ID %in% TRBJ1)
```

## bring in DESeq2 data from Wylie et al. (2026) and combine with filtered PBMC datasets for combined figure

```{r}
# files generated from the supplementary file for plasma and PBMC (Wylie et al. (2026))
PBMC_DESeq2_IBC_Gen = read_delim(file = "IBC_PBMC_DESeq2_normalized_counts.tsv", delim = "\t") %>% 
  pivot_longer(cols = PBMCH01:PBMCI10, names_to = "DESeq2_sample_name", values_to = "DESeq2-normalized counts")

Plasma_DESeq2_IBC_Gen = read_delim(file = "IBC_Plasma_DESeq2_normalized_counts.tsv", delim = "\t") %>% 
  pivot_longer(cols = PlasmaH01:PlasmaI10, names_to = "DESeq2_sample_name", values_to = "DESeq2-normalized counts")

# file was generated in 2024-08-14_DESeq2_IBC_30nt_cutoff.Rmd
PBMC30nt_DESeq2_IBC_Gen = read_delim(file = "IBC_PBMC_filtered30nt_DESeq2_normalized_counts.tsv", delim = "\t") %>% 
  select(-...1) %>% # get rid of rownames
  pivot_longer(cols = PBMCI01:PBMCH13, names_to = "DESeq2_sample_name", values_to = "DESeq2-normalized counts") %>% 
  left_join(DESeq2_IBC_metadata) %>% 
  mutate("Read Lengths" = "<=30 nt filtered read lengths")

combo_DESeq2_IBC_Gen = rbind(PBMC_DESeq2_IBC_Gen, Plasma_DESeq2_IBC_Gen) %>%   
  left_join(DESeq2_IBC_metadata) %>% 
  mutate("Read Lengths" = "All read lengths") %>% 
  select(-Type)

combo_DESeq2_IBC_Gen2 = rbind(combo_DESeq2_IBC_Gen, PBMC30nt_DESeq2_IBC_Gen) %>% 
  mutate("Sample and filtering" = paste0(Sample.Type," ",`Read Lengths`)) %>% 
  filter(ID %in% c("TRBJ1_1","TRBJ1-1","TRBJ1_2","TRBJ1-2","TRBJ1_6","TRBJ1-6")) %>% 
  mutate(Target = case_when(ID %in% c("TRBJ1_1","TRBJ1-1") ~ "TRBJ1-1",
                            ID %in% c("TRBJ1_2","TRBJ1-2") ~ "TRBJ1-2",
                            ID %in% c("TRBJ1_6","TRBJ1-6") ~ "TRBJ1-6"))

combo_DESeq2_IBC_Gen2$`Sample and filtering` = factor(combo_DESeq2_IBC_Gen2$`Sample and filtering`, levels = c("PBMC All read lengths","PBMC <=30 nt filtered read lengths","Plasma All read lengths"))

combo_DESeq2_IBC_Gen2$Disease.category = factor(combo_DESeq2_IBC_Gen2$Disease.category, levels = c("Healthy","non-IBC", "IBC"))

```


## Final Figure1C
```{r}
combo_DESeq2_IBC_Gen2 %>% 
  ggplot(aes(x = Disease.category, y = `DESeq2-normalized counts`, fill = Disease.category))+
  geom_violin(aes(alpha = 0.5))+
  geom_dotplot(binaxis='y', stackdir='center', dotsize=1)+
  facet_grid(Target~`Sample and filtering`)+
  scale_y_log10()+
  theme_bw()+
  theme(axis.title.x=element_blank(),legend.position="none")+
  scale_fill_manual(values = c("#3b3b3b","#4173C6","#FF5733"))

```
## Final Figure S2
```{r}

combo_DESeq2_IBC_Gen2$Sample = factor(combo_DESeq2_IBC_Gen2$Sample, levels = c("H01", "H02", "H03", "H04", "H05", "H06", "H07", "H08", "H09", "H10", "H11", "H12", "H13", "B01", "B02", "B03", "B04", "B05", "B06", "I01", "I02", "I03", "I04", "I05", "I06", "I07", "I08", "I09", "I10"))

combo_DESeq2_IBC_Gen2 %>%
  ggplot(aes(x = Sample, y = `DESeq2-normalized counts`, color = Target, ID=DESeq2_sample_name))+
  geom_point(shape = 1, position = position_dodge())+
  # theme(axis.text.x=element_text(angle=90,hjust = 1))+
  facet_grid(`Sample and filtering`~., scales = "free_y")+
  scale_y_log10()+
  theme_bw()+
  labs(y= "DESeq2-normalized counts")+
  theme(axis.text.x=element_text(angle=90,hjust = 1))+
  scale_color_manual(values = c("purple","darkorange","darkred"))+
  theme(axis.title.x=element_blank(),legend.position="top")
```







